sdmTMB statistical model description

### 2021-12-02 doc/appendix-model.Rmd
Cross off some TODOs in the model appendix

* Get into smoothers, breakpoints, etc. here

@@

@@

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.

### Random walk spatiotemporal random fields

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

## Spatially varying coefficients

@@

>>>>

@@

Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to `mgcv` users fitting GAMs (generalized additive models). Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. Within these smooths, the same syntax commonly used in mgcv::s() can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, `+ s(month, bs = "cc", k = 12)`.

While `mgcv` can fit unpenalized (e.g. B-splines, and others) or penalized splines (P-splines), sdmTMB only uses penalized splines. 
The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed effects modeling framework. This is a similar approach implemented in `gamm4` and `brms`, among other packages.

@@

The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses. This function consists of two pieces, so that for $x < b_{1}$

$$s(x) = x \cdot b_{0}$$
and for $x > b_{1}$

$$s(x) = b_{1} \cdot b_{0}$$
In both cases, $b_{0}$ represents the slope of the function up to a threshold, and the product $b_{1} \cdot b_{0}$ represents the value at the asymptote. No constraints are placed on parameters $b_{0}$ or $b_{1}$.

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance. The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is generally expected to be $-1 < \rho < 1$.

### Random walk spatiotemporal random fields (RW)

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.

@@

## Spatially varying coefficients (SVC)

@@

where $\mu$ and $\phi$ are the mean and standard deviation of the natural log of the random variable. If $X$ is lognormally distributed, $E[log(X)] = \mu$ and $Var(log(X)) = \phi$. The expectation of $X$ is $exp(u + \frac{\phi^2}{2})$ and variance of X is $\left(exp(\phi^2) - 1 \right)exp(2\mu + \phi^2)$.
